\documentclass[a4paper,12pt]{scrartcl}

\usepackage[utf8]{inputenc} 
\usepackage[T1]{fontenc}
\usepackage[english]{babel}
\usepackage{amsmath, amssymb,amsfonts}
\usepackage{graphicx}
\usepackage{csquotes}
\usepackage{geometry}
\usepackage{float}
\usepackage{framed, xcolor} 
\usepackage{scrlayer-scrpage}
\usepackage{siunitx}
\usepackage{subcaption}
\usepackage{chemgreek}
\usepackage{chemformula}
\usepackage[bookmarks,colorlinks=true]{hyperref}
\usepackage{booktabs}

% 1. ZUERST DIE VARIABLEN DEFINIEREN
\newcommand{\VERSUCHSDATUM}{March 2, 2026}
\newcommand{\PROTOKOLLDATUM}{\today}
\newcommand{\VerfasserEINS}{Julian Brügger}
\newcommand{\MatNoEINS}{3715444}
\newcommand{\EmailEINS}{st190050@stud.uni-stuttgart.de}
\newcommand{\StudiengangEINS}{B.Sc. Chemie}
\newcommand{\VerfasserZWEI}{Benedict Roßkopf}
\newcommand{\MatNoZWEI}{3718292}
\newcommand{\EmailZWEI}{st188124@stud.uni-stuttgart.de}
\newcommand{\StudiengangZWEI}{B.Sc. Simulation Technology}
\newcommand{\BETREUER}{}
\newcommand{\GRUPPENNR}{Gruppe X} % Darf nicht leer sein, wenn im Header genutzt
\newcommand{\VERSUCHSNR}{1}
\newcommand{\VERSUCHSNAME}{Reaction enthalpies and harmonic vibrational frequencies}

% 2. DANN DAS LAYOUT
\geometry{left = 2.5cm, top = 3cm}
\renewcaptionname{english}{\figurename}{Fig.}
\renewcaptionname{english}{\tablename}{Tab.}

\sisetup{
    detect-weight=true, 
    detect-family=true,
    locale=UK,
    exponent-product = \cdot,
    separate-uncertainty=true,
    per-mode = symbol-or-fraction
}

\DeclareSIUnit{\angstrom}{\text{\AA}}

\hypersetup{
    colorlinks,
    linktocpage,
    citecolor=black,
    filecolor=black,
    linkcolor=black,
    urlcolor=black
}

\setlength{\parindent}{0 mm}
\setlength{\parskip}{2 mm} 

\pagestyle{scrheadings}
\clearpairofpagestyles % Löscht alte Voreinstellungen
\ihead{\VERSUCHSNR} 
\ofoot{\thepage} 

\begin{document}

\begin{titlepage}
\begin{center}
\Huge{\textbf{\VERSUCHSNAME}}\\
\vspace{10mm}
\Large{Protocol for the Computational Chemistry course by \\ \textbf{\VerfasserEINS\;\& \VerfasserZWEI }}\\
\vspace{10mm} 
\Large{University of Stuttgart}\\
\end{center}
\begin{center}
\begin{tabular}{ll}
\large{authors:}        & \large{\VerfasserEINS, \MatNoEINS} \\
                        & \large{\EmailEINS} \\
                        \vspace{2mm}\\
                        & \large{\VerfasserZWEI, \MatNoZWEI} \\
                        & \large{\EmailZWEI} \\						
                        \vspace{0cm}\\
\large{date of experiment:}	& \large{\VERSUCHSDATUM} \\
\vspace{0cm}\\
\large{submission date:} & \large{\PROTOKOLLDATUM}
\end{tabular}
\end{center}
\end{titlepage}

\section{Introduction}

The objective of this computational study is to evaluate the reaction enthalpies and harmonic vibrational frequencies of small molecules, specifically focusing on the hydrogenation of carbon monoxide and formaldehyde. Standard quantum chemical calculations typically yield results for states in a vacuum at 0 K, which does not reflect experimental conditions; therefore, corrections such as the addition of vibrational energy are necessary to improve the results and provide accurate predictions. The reaction enthalpy at 298 K is determined by the following relation:
\begin{equation}
\Delta H_r(298 K) = \Delta E_0 + \Delta E_{ZPE} + \Delta E(0 K \rightarrow 298 K) + \Delta N k_B \cdot 298 K
\label{eq1}
\end{equation}
This combines the electronic energy with the zero point vibrational energy, thermal corrections, and the pressure-volume work. To account for electron correlation effects neglected in Hartree-Fock theory, second-order Møller-Plesset perturbation theory (MP2) is employed, where the total energy is defined as:
\begin{equation}
E_{MP2} = E_{HF} + E_2
\end{equation}
A central part of the analysis involves harmonic vibrational frequencies, derived by approximating the potential energy surface around a stationary point using a Taylor series truncated after the quadratic term. By transforming to mass-weighted coordinates $q_k = \sqrt{M_k} \Delta r_k$, the Hamiltonian decouples into independent 1D harmonic oscillators, allowing the calculation of energy eigenvalues:
\begin{equation}
E_{vib} = \sum \hbar \omega_i \left(v_i + \frac{1}{2}\right)
\end{equation}
The resulting zero point energy is given by:
\begin{equation}
E_{ZPE} = \frac{1}{2} \sum \hbar \omega_i
\end{equation}
While this harmonic approximation is computationally feasible, it is primarily valid for low-lying vibrational states and serves as a tool for verifying stationary points, where a transition state is identified by exactly one imaginary frequency. By comparing these theoretical frequencies with experimental data, the performance of various methods such as DF-RHF, DF-MP2, and DF-RKS/B3LYP using the def2-TZVPP basis set can be assessed.

\section{Results}

\subsection{Reaction enthalpies}

The reaction enthalpies of the hydrogenation of carbonmonoxide and the hydrogenation of its product were calculated using the DF-MP2/def2-TZVPP base set and are listed in \autoref{tab1}.

% Table 1: Reaction Enthalpies
\begin{table}[H]
\centering
\caption{Reaction enthalpies $\Delta H_r(298\text{ K})$ in kJ/mol calculated using MP2/def2-TZVPP.}
\label{tab1}
\begin{tabular}{lcc}
\toprule
Reaction & MP2/def2-TZVPP & Experiment \\ \midrule
 \ch{CO + H2 -> H2CO} & 3.93 & 1.92 \\
\ch{H2CO + H2 -> H3COH} & $-90.27$ & $-92.55$ \\ \bottomrule
\end{tabular}
\end{table}

The point groups of the investigated molecules were determined using the Schoenflies notation as $D_{\infty h}$ for carbon monoxide (CO) and hydrogen (\ch{H2}), $C_{2v}$ for formaldehyde (\ch{H2CO}), and $C_s$ for methanol (\ch{H3COH}).

The number of vibrational degrees of freedom was determined using the formula $3N-5$ for linear molecules and $3N-6$ for non-linear molecules. For the linear molecules carbon monoxide (CO) and hydrogen (\ch{H2}), the calculation yields 1 vibrational degree of freedom each. For the non-linear molecules, the formaldehyde molecule (\ch{H2CO}) possesses 6 vibrational degrees of freedom, which are classified into stretching vibrations (symmetric and asymmetric) and bending vibrations (scissoring, rocking, wagging, and twisting), while methanol (\ch{H3COH}) has 12 vibrational degrees of freedom.

The three energy-termns $\Delta E_0$, $\Delta E_{\text{ZPE}}$  und $\Delta E(0 K \rightarrow 298 K)$ in \autoref{eq1} represent the internal energy $U$, the $\Delta N \cdot k_B \cdot T$ term represents the volume work. The $\Delta E_0$ term stands for the differnence of the energy niveaus of the electrons at 0 K. The $\Delta E_{\text{ZPE}}$ term stands for the energy of the vibration modes at said temperature and the $E(0 K \rightarrow 298 K)$ represents the energy increase of the translation, rotation and vibration modes going from 0 to 298 K.

The primary approximations within the terms of Equation 1 include the Born-Oppenheimer approximation, which separates electronic and nuclear motion to calculate $\Delta E_{el}$, and the harmonic approximation, which treats molecular vibrations as simple harmonic oscillators for the determination of $\Delta E(ZPE)$ and the vibrational contributions to the thermal energy. Additionally, the thermal correction $\Delta E(0 \text{ K} \rightarrow 298 \text{ K})$ typically relies on the ideal gas law and the rigid rotor model to simplify the translational and rotational energy contributions.

The calculated values of the vibrational frequencies, using DF-RHF, DF-MP2, and DF-RKS/B3LYP with a /def2-TZVPP basis set are displayed in \autoref{tab2}.

% Table 2: Harmonic Vibrational Frequencies
\begin{table}[H]
\centering
\caption{Harmonic vibrational frequencies of $\text{H}_2\text{CO}$ using various electronic structure methods and the def2-TZVPP basis set. All frequencies are given in $\text{cm}^{-1}$.}
\label{tab2}
\begin{tabular}{cccccc}
\toprule
Mode & Symmetry & HF & MP2 & B3LYP & Experiment \\ \midrule
ZPE & $(A_1)$ & 0.815 eV & 0.737 eV & 0.738 eV & \\ \midrule
1 & $B_1$ & 1351.4 & 1206.8 & 1187.4 & 1167.3 \\
2 & $B_2$ & 1360.5 & 1278.0 & 1251.8 & 1249.1 \\
3 & $A_1$ & 1617.0 & 1553.2 & 1519.3 & 1500.1 \\
4 & $A_1$ & 1818.8 & 1764.0 & 1754.7 & 1746.1 \\
5 & $A_1$ & 3028.8 & 2976.6 & 2940.2 & 2782.5 \\
6 & $B_2$ & 3108.3 & 3051.3 & 3006.7 & 2843.1 \\ \bottomrule
\end{tabular}
\end{table}

The geometry must be optimized prior to a harmonic frequency calculation because the harmonic approximation is only valid at a stationary point on the potential energy surface. At such a point, the first derivative of the energy with respect to nuclear coordinates (the gradient) is zero. If a frequency calculation yields imaginary frequencies, it indicates that the structure is not at a local minimum but rather at a saddle point, which represents a transition state or an unstable configuration.

While the Third Law of Thermodynamics states that $0\text{ K}$ cannot be reached, the zero-point energy is a fundamental quantum mechanical consequence of the Heisenberg Uncertainty Principle, ensuring that even at a theoretical absolute zero, a molecule retains a minimum vibrational energy of $\frac{1}{2} h\nu$ per mode.


\end{document}